Model-agnostic Interpretable Machine Learning

### 2021-08-21 2021-bips-practice-phd-defence/2021-bips-practice-phd-defence.Rmd
use xaringan

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output: ioslides_presentation
---

@@

## Outline

@@

@@

>- 1800s: Linear regression models
>- 1960s: Rule-based ML
>- 2001:  Random Forest
>- 2012:  Deep Learning Boom kicked off
>- 2016:  Interpretable ML / Explainable AI boom

@@

```{r}
load(url("https://github.com/christophM/interpretable-ml-book/blob/master/data/bike.RData?raw=true"))
library("mlr")
library("iml")
library("ggplot2")
library(dplyr)
library(tidyr)

@@

## Step 1: Define grid over feature
grid.dat = data.frame(temp = unique(res.raw$temp), .id = 1, temp2 = 1)
p = ggplot(grid.dat) + geom_vline(aes(xintercept = temp)) +
  scale_y_continuous("Prediction", limits = c(min(res.raw$.y.hat), max(res.raw$.y.hat))) +
  ggtitle("1) Define grid along feature") +
  scale_x_continuous("Feature")
  #scale_y_continuous("Prediction")
p
```

@@

```{r}
## Step 2: Get predictions for data set when at grid
grid.dat = data.frame(temp = unique(res.raw$temp))
res.raw$tt = res.raw$temp
p = ggplot(res.raw, aes(x = temp, y = .value)) +
  geom_vline(data = grid.dat, aes(xintercept = temp)) +
  geom_point(aes(color = factor(.id)), size = 3)+
  scale_color_discrete(guide = "none") +
  ggtitle("2) Model predictions at grid points") +
  scale_x_continuous("Feature") +
  scale_y_continuous("Prediction", breaks = NULL)
p
```

@@

```{r}
res.raw$.id2 = factor(res.raw$.id)
p = ggplot(res.raw, aes(x = temp, y = .value))  +
  #geom_vline(aes(xintercept = temp), data = grid.dat) +
  geom_line(aes(group = .id), size  = 1)  +
  geom_point(data = res.raw[c("temp", ".value", ".id2")], aes(color = .id2), size = 3) +
  scale_color_discrete(guide = "none") +
  ggtitle("3) Line per data instance -> ICE curves") +
  scale_x_continuous("Feature") +
  scale_y_continuous("Prediction", breaks = NULL)
p
```

@@

@@

```{r}
p
```

```{r}
res = res.raw %>% group_by(temp) %>%
  summarize(pdp = mean(.value))

@@

# Difference between model-agnostic and model-specific interpetation

@@

## Most Important

@@

# Paper: Subgroups

## Problem: Dependent features

Show example with dependent features for PFI

## Solution: Conditional variants

Show solution

* But: Not meaningful for PDP, or other measure more useful (LOESS)
* But: Conditioning unclear
* But: Intepretation difficult

## Our solution: Permute in subgroups

TODO: Show algorithm

## Some results from our paper

TODO: Show results or application

# Paper: IML Inference

## Problem: Uncertainty quantification

PDP and PFI and most other just descriptive tools
Often practictioner draw confidence intervals
but only based on monte carlo approximation

* But what is even the ground truth?
* And what about the uncertainty in the model fit?
* How could we estimate the uncertainty?

## Solution: Treat as statistical estimator

## How to Compute the variance

## Problem of too low variance estimates

# Paper: Quantifying

TODO: Copy as much as possible from the talk that already exists for this paper

Motivation: want to quantify in a model-agnostic way how complex a model is

## Functional Decomposition

Explain how functional decomposition works

## ALE

Formula:

TODO: For estimation show the images

## The three measures

* Mean effect complexity
* non linearity
* number of features

## Using in benchmarking (multi-objective)

@@

@@

- For inference, practitioner rely on interpretable models
- remember: interpretable models have parameters that map to data-generating process
- but: situations where more flexible models match DGP better (measured in test loss)
- we have: model-agnostic interpretation
- But: only describes the model behavior
- Question: Can we link PD and PFI to the DGP? and even do inference?

## Example

TODO: Show PDP and PFI from some papers?

@@

TODO: simple image

@@

## Show simulation results

# Thank You for Your Attention

>>>>

---
title: "Model-agnostic Interpretable Machine Learning"
author: "Christoph Molnar"
date: "2021-08"
institute: "BIPS bremen"
output:
  xaringan::moon_reader
---

@@

# Outline

@@

---

@@

Interpretable ML has 200 year old roots, but consolidation only started in 2016.
--

- 1800s: Linear regression models

--

- 1960s: Rule-based ML

--

- 2001:  Random Forest

--
- 2012:  Deep Learning Boom kicked off

--

- 2016:  Interpretable ML / Explainable AI boom

???

Now impress audience with history

---

@@

--

```{r, out.width = 500}
load(url("https://github.com/christophM/interpretable-ml-book/blob/master/data/bike.RData?raw=true"))
library("mlr")
library("iml")
library("ggplot2")
library(dplyr)
library(tidyr)

@@

## Step 1: Define grid over feature
grid.dat = data.frame(temp = unique(res.raw$temp), .id = 1, temp2 = 1)
p = ggplot(grid.dat) + geom_vline(aes(xintercept = temp)) +
  scale_y_continuous("Prediction", limits = c(min(res.raw$.y.hat), max(res.raw$.y.hat))) +
  ggtitle("1) Define grid along feature") +
  scale_x_continuous("Feature")
  #scale_y_continuous("Prediction")
p
```
---

@@

```{r, out.width = 500}
## Step 2: Get predictions for data set when at grid
grid.dat = data.frame(temp = unique(res.raw$temp))
res.raw$tt = res.raw$temp
p = ggplot(res.raw, aes(x = temp, y = .value)) +
  geom_vline(data = grid.dat, aes(xintercept = temp)) +
  geom_point(aes(color = factor(.id)), size = 3)+
  scale_color_discrete(guide = "none") +
  ggtitle("2) Model predictions at grid points") +
  scale_x_continuous("Feature") +
  scale_y_continuous("Prediction", breaks = NULL)
p
```

---

@@

```{r, out.width = 500}
res.raw$.id2 = factor(res.raw$.id)
p = ggplot(res.raw, aes(x = temp, y = .value))  +
  #geom_vline(aes(xintercept = temp), data = grid.dat) +
  geom_line(aes(group = .id), size  = 1)  +
  geom_point(data = res.raw[c("temp", ".value", ".id2")], aes(color = .id2), size = 3) +
  scale_color_discrete(guide = "none") +
  ggtitle("3) Line per data instance -> ICE curves") +
  scale_x_continuous("Feature") +
  scale_y_continuous("Prediction", breaks = NULL)
p
```
---

@@

--

```{r, out.width = 500}
p
```

---

@@

```{r, out.width = 500}
res = res.raw %>% group_by(temp) %>%
  summarize(pdp = mean(.value))

@@

---

## Difference between model-agnostic and model-specific interpetation

@@

---

## Most Impactful Work

```{r, out.width = 300}
knitr::include_graphics("../images/book.jpg")
```

---

@@

---

@@

---

## Goal of Paper

Treat permutation feature importance (PFI) and partial dependence plots (PDP) as statistical estimators.

---

## Statistical Modeling

- Assumptions made about the data-generating process
- Mapping between model parameters and properties of data-generating process
- Statistical Inference possible

TODO: Draw image which shows link between model and DGP

---

@@

- For inference, practitioner rely on interpretable models
- But: situations where more flexible models match DGP better (measured in test loss)
- Interpretation possible with PDP and PFI (among others)
- However only describes the model behavior
- Question: Can we link PD and PFI to the DGP? and even do inference?

---

@@

TODO: simple image with DGP + model. External tools, PDP and PFI, applied externally and compared.

---

@@

---

## Thank You for Your Attention

```{r, out.width = 500}
knitr::include_graphics("../images/applause.gif")
```
